\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-10-17 eg6-letters/NIM/EG6_RTPC_NIM.tex
With line numbers for NIM submission
added the answers for the last 25 questions

\documentclass[preprint,5p]{elsarticle}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}  
\usepackage{amssymb}  
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor=blue, citecolor=blue}
\usepackage[displaymath]{lineno}
\begin{document}

@@

@@

Until recently, the Thomas Jefferson National Accelerator Facility, in 
Newport News, Virginia, USA, has provided high power electron beams of 
up to 6 GeV energy and 100$\%$ duty factor to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{CLASref}, located
in Hall-B, was based on a superconducting toroidal magnet and composed of 
several sub-detectors. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
       particles~\cite{DCref}.
 \item Superconducting toroidal magnet to bend the trajectories 
       of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Threshold Cherenkov Counters (CC) for electron identification at momenta 
       $<2.7$ GeV~\cite{CCref}.
 \item Scintillation Counters (SC) to identify charged hadrons by measuring their 
       time of flight~\cite{TOFref}.
 \item Electromagnetic Calorimeters (EC) for identification of electrons, 
       photons and neutrons~\cite{ECref}.
\end{itemize}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nuclei) necessitates an upgrade of the photon detection system. 
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 
experiment in 2005~\cite{FX}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid magnet was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on Helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in the RTPC. The CLAS detector supplemented with both IC and RTPC was used 
during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 
2005~\cite{BONUS-NIM}. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib} and the tracking algorithm in section
\ref{sec_tracking}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230~mm diameter shell and cryostat 
wall of the solenoid magnet bore.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of cylindrical gas electron multipliers (GEM) and 
detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves with 
independent GEM amplification systems that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
  \item The 6-atm $^4$He target extends along the beam axis; it 
      is a 284-mm-long, 6-mm-diameter Kapton straw with a 27-$\mu$m wall.
   \item The first gas gap covers the radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at $- 4260$~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.70]{GEM_photo.jpg}
\caption{Image of a typical GEM foil similar to the one used for our RTPC 
\cite{GEM_ref_pic}.} 
   \label{fig:GEMs}
\end{figure}

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low spark rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 300~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers\footnote{The GEM foils were produced by 
Tech-Etch, Inc.}. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% Ne-DME mixture. This choice 
has been made in order to balance the energy deposit, which is critical
for proper particle identification, with a reasonable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the axial 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture.

One of the important challenge in developing the radial TPC was to obtain a 
good support structure for the GEM foils to allow a tractable 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each end of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the upstream end of the cylinder 
was fixed to the main mechanical support structure. This design only left a 
light fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system had 3200 readout pads. These elements were
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} illustrates the configuration of the 5 by 4.45 mm pads,
where the shift between the rows was implemented to reduce aliasing. Each half of the 
RTPC had 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads were grouped to 16 channels pre-amplifier boards. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6-m long ribbon 
cable that connects to it.

@@

The readout system, similar to the original BoNuS RTPC system~\cite{BONUS-NIM}, 
was based on the front end electronic boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. 
The Front End Card (FEC) 
hosted the analogue receiver circuit and the subsequent amplification and 
digitization stage, based on the PASA~\cite{ALICE-PASA} and ALTRO~\cite{ALICE-ALTRO} 
ASICS. A Readout Control Unit (RCU) board was used to distribute the trigger signal 
to FECs and for data readout, with each RCU handling up to 25 FECs. Communication 
between readout controller application (ROC) and FECs was performed trough a custom 
backplane, implementing a low-voltage signal bus. The RCU communicated with the 
CLAS DAQ system trough a 200 MB/s optical link, connected to a data acquisition 
PC hosting a ReadOut Receiver Card (RORC). This PC hosted the ROC used to interface 
with the CLAS DAQ system. An Ethernet link was also present, for slow-controls and 
monitoring.

@@

The trigger signal also initiates RCU readout operation from FEC boards. All the 
measured samples from active channels are reported, together with a channel 
identifier and a timestamp, to the ROC application, that in turns sends them on 
the main event builder. Readout from FECs occurs in ``block-transfer'' mode, 
making use of the ALTRO internal multi-event buffer. This feature could be 
exploited thanks to the new RCU boards developed for the RTPC readout 
system\footnote{In the original BONUS system, the U2F readout controller had 
no multi-event capability and used a USB connection rather than fiber 
optics for readout. These factors limited the maximum readout rate from the 
BONUS detector to about 500 Hz.}. In order to read all the detector readout 
pads, four FEC crates were used, each equipped with 6 boards, plus a ROC. A 
schematic of the readout system, for a single crate, is reported in 
figure~\ref{fig:ReadoutScheme}. This configuration permitted to reduce the dead 
time associated to FEC readout operations, which is scaling linearly with the number of 
boards in the crate. During the 2009 run, the system was successfully operated 
with a DAQ rate of 3.1 kHz and a live time of $70 \%$, for a luminosity of 
about $10^{34}$~cm$^{-2}$.s$^{-1}$ and a beam energy of 6.067 GeV.

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100 ns.

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, with each time an increasing number of tracks 
reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, the cylindrical symmetry insures that this maximum drift distance is the same for all 
tracks. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can therefore infer the drift 
speed of the electrons in the RTPC.\\

@@

To measure the maximum drift time, we used the time profile of hits from identified tracks 
shown in Fig.~\ref{fig:TDC_profile}. We can clearly observe the 
dropping edge expected from geometrical considerations. We define a value $T_{Max/2}$ at 
which the dropping edge passes half the maximum number of hits in the 
histogram. This value was measured in bins along the 200~mm RTPC's length to take into 
account variations in the electric and magnetic field in the RTPC (see 
Fig.~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

@@

Due to the non perfect experimental conditions, in particular possible contamination 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around run number 
61600. While our gas system was kept slightly over 
atmospheric pressure to limit contamination from air or other external gases, it 
is likely that this leak was the source of modification of the drift speed.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe significant change in 
the drift speed before and after run 61600, while variations within these periods 
are ~2\%.

@@

In summary, we obtain from our calibration a parame\-tri\-zation of the drift speed 
as a function of both position along the beam axis and the run number. These 
functions were extracted for our entire data set and implemented in the 
track reconstructions code.

\subsection{Drift Paths Calibration}

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used this calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometrical accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) was extracted as a function of time. 
In Fig.~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift path is easily identified and eventually fitted for 
implementation in our reconstruction codes.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Fig.~\ref{fig:Drift_run_number_1}). Interestingly, we found very similar drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

To calibrate the gains, we compared the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charge along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains were refined using correction factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compared it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Fig.~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is similar to those of real tracks. About 
18\% of the readout channels exhibit large contributions from this noise 
characteristic. Due to its unique time-energy correlation for the given 
channels, the noise could be removed on an event by event and channel by 
channel basis without significant loss of good signals. The result of the
procedure is illustrated in the bottom panel of Fig.~\ref{fig:noise}.

@@

The primary data sample used for calibration and performance assessment of the
RTPC was elastic scattering with a 1.2 GeV electron beam. The electron's momentum
and direction is measured with CLAS, which uniquely determines the expected recoiling
$^4$He kinematics. Matching requirements between reconstructed and expected $z$-vertex
and direction of the RTPC track provides a clean selection of elastically-scattered
$^4$He, shown in Fig.~\ref{fig:w}.

\begin{figure}[tb]\centering
  \includegraphics[width=8cm]{fit_W_distribution_l.png}
  \caption{The $W$ distribution calculated from electron kinematics before (``inclusive'')
  and after (``exclusive'') requiring a matching track in the RTPC.\label{fig:w}}
\end{figure}

>>>>

\documentclass[preprint,5p]{elsarticle}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}  
\usepackage{amssymb}  
\usepackage{hyperref}
\providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}
\hypersetup{colorlinks=true, urlcolor=blue, citecolor=blue}
\usepackage[displaymath]{lineno}
\linenumbers
\begin{document}

@@

%\begin{keywords}
%\textcolor{red}{Raphael please add the list you provided originally},
%gas electron multipliers.
%\end{keywords}

@@

Until recently, the Thomas Jefferson National Accelerator Facility, in 
Newport News, Virginia, USA, has provided high power electron beams of 
up to 6 GeV energy and 100$\%$ duty factor to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{Mecking:2003zu}, located
in Hall-B, was based on a superconducting toroidal magnet and composed of 
several sub-detectors. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
       particles~\cite{Mestayer:2000we}.
 \item Superconducting toroidal magnet to bend the trajectories 
       of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Threshold Cherenkov Counters (CC) for electron identification at momenta 
    $<2.7$ GeV/c~\cite{Adams:2001kk}.
 \item Scintillation Counters (SC) to identify charged hadrons by measuring their 
       time of flight~\cite{Smith:1999ii}.
 \item Electromagnetic Calorimeters (EC) for identification of electrons, 
       photons and neutrons~\cite{Amarian:2001zs}.
\end{itemize}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nucleus) necessitates an upgrade of the photon detection system.  
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 experiment 
in 2005~\cite{Girod:2007aa}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the CLAS detector and the IC from the 
large flux of low energy M{\o}ller electrons, a 5~T solenoid magnet was placed 
around the target to shield the detectors.  To detect recoiling $\alpha$ 
particles from the coherent DVCS on Helium, a new radial time projection 
chamber (RTPC) was developed to track low energy nuclear fragments. The 
solenoid field was used to bend tracks and measure momentum of particles in the 
RTPC. The CLAS detector supplemented with both IC and RTPC was used during a 
three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 
2005~\cite{Fenker:2008zz}. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib} and the tracking algorithm in section
\ref{sec_tracking}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from 
coherent DVCS have an average momentum around 300~MeV/c (12~MeV kinetic 
energy). Such low energy $\alpha$ particles are stopped very rapidly, so the 
RTPC was designed to be as close as possible to the target and fit inside the 
230~mm diameter shell and cryostat wall of the solenoid magnet bore.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of cylindrical gas electron multipliers (GEM) and 
detected by the readout system on the external shell of the detector as 
illustrated in Figure~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves 
with independent GEM amplification systems that cover about 80\% of the 
azimuthal angle.

We detail here the different regions shown in Figure~\ref{fig:RTPC_1_4} 
starting from the beam line towards larger radius:\\
\begin{itemize}
  \item The 6~atm $^4$He target extends along the beam axis; it is a 284~mm 
     long, 6~mm diameter Kapton straw with a 27~$\mu$m wall.
   \item The first gas gap covers the radial range from 3~mm to 20~mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gas gap region extends between 20~mm and 30~mm and is 
      filled with the gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether 
      (DME).  This region is surrounded by a 4~$\mu$m thick window made of 
      aluminized Mylar set at $-4260$~V to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and 
      extends from the cathode to the first GEM, 60~mm away from the beam axis.  
      The average electric field in this region is perpendicular to the beam 
      and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66~mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600~V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charges 
      after they have been multiplied by the GEMs. Pre-amplifiers are plugged 
      directly on its outer side and transmit the signal to the data 
      acquisition electronics.
\end{itemize}

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.70]{GEM_photo.jpg}
\caption{Image of a typical GEM foil similar to the one used for our RTPC 
\cite{Sauli:2016eeu}.} 
   \label{fig:GEMs}
\end{figure}

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low spark rate~\cite{Beringer:1900zz}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 300~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers\footnote{The GEM foils were produced by 
Tech-Etch, Inc.}. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Figure~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong electric 
field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% Ne-DME mixture. This choice 
has been made in order to balance the energy deposit, which is critical
for proper particle identification, with a reasonable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{Biagi:1999nwa} 
showed that with the 5~T solenoidal magnetic field, we would have a Lorentz 
angle of about 23$^\circ$ with this gas mixture.

One of the important challenges in developing the RTPC was to obtain a good 
support structure for the GEM foils to allow a tractable installation of the 
GEMs. At the same time, we wanted to keep the material budget small in the 
forward region where we detect other particles in subsequent detectors. We 
successfully realized these prerequisites by using fiber glass rings glued to 
each end of the GEM foils to form self supporting cylinders that could be 
installed independently in the RTPC after gluing and soldering operations. The 
rigidity of the GEM foils was enough
for the structure to be self-supporting and only the upstream end of the cylinder 
was fixed to the main mechanical support structure. This design only left a 
light fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system had 3200 readout pads. These elements were
located at the end of the amplification region, 69~mm from the central axis.
Figure \ref{fig:PADs} illustrates the configuration of the 5 by 4.45~mm pads,
where the shift between the rows was implemented to reduce aliasing. Each half 
of the RTPC had 40 rows and 40 columns of pads. The shaded region in 
Figure~\ref{fig:PADs} shows how pads were grouped to 16 channels pre-amplifier 
boards. The pre-amplifier boards, already employed in the BoNuS 
RTPC~\cite{Fenker:2008zz}, serve the dual purpose of inverting the RTPC signals 
polarity -- from negative to positive -- to match the requirements of the 
subsequent readout system, and driving the 6~m long ribbon cable that connects 
to it.

@@

The readout system, similar to the original BoNuS RTPC system~\cite{Fenker:2008zz}, 
was based on the front end electronic boards originally developed for the ALICE 
TPC readout system at CERN~\cite{ALICE-FEE}. The Front End Card (FEC) hosted 
the analogue receiver circuit and the subsequent amplification and digitization 
stages, based on the PASA~\cite{Soltveit:2012jp} and 
ALTRO~\cite{EsteveBosch:2003bj} ASICS. A Readout Control Unit (RCU) board was 
used to distribute the trigger signal to FECs and for data readout, with each 
RCU handling up to 25 FECs. Communication between readout controller 
application (ROC) and FECs was performed trough a custom backplane, 
implementing a low-voltage signal bus. The RCU communicated with the CLAS DAQ 
system trough a 200 MB/s optical link, connected to a data acquisition PC 
hosting a readout receiver card. This PC hosted the ROC used to interface with 
the CLAS DAQ system. An Ethernet link was also present, for slow-controls and 
monitoring.

@@

The trigger signal also initiates RCU readout operation from FEC boards. All the 
measured samples from active channels are reported, together with a channel 
identifier and a timestamp, to the ROC application, that in turns sends them on 
the main event builder. Readout from FECs occurs in ``block-transfer'' mode, 
making use of the ALTRO internal multi-event buffer. This feature could be 
exploited thanks to the new RCU boards developed for the RTPC readout 
system\footnote{In the original BoNuS system, the U2F readout controller had no 
multi-event capability and used a USB connection rather than fiber optics for 
readout. These factors limited the maximum readout rate from the BoNuS detector 
to about 500 Hz.}. In order to read all the detector readout pads, four FEC 
crates were used, each equipped with 6 boards, plus a ROC. A schematic of the 
readout system, for a single crate, is reported in 
Figure~\ref{fig:ReadoutScheme}. This configuration permitted to reduce the dead 
time associated to FEC readout operations, which is scaling linearly with the 
number of boards in the crate. During the 2009 run, the system was successfully 
operated with a DAQ rate of 3.1 kHz and a live time of $70 \%$, for a 
luminosity of about $10^{34}$~cm$^{-2}\cdot$s$^{-1}$ and a beam energy of 
6.064~GeV.

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100~ns.

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{Biagi:1999nwa} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, with each time an increasing number of tracks 
reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, the cylindrical symmetry insures that this maximum drift distance is the 
same for all tracks. We illustrate this in Figure~\ref{fig:RTPC_signals}, where 
we depict a typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), 
we can therefore infer the drift speed of the electrons in the RTPC.\\

@@

To measure the maximum drift time, we used the time profile of hits from 
identified tracks shown in Figure~\ref{fig:TDC_profile}. We can clearly observe 
the dropping edge expected from geometrical considerations. We define a value 
$T_{Max/2}$ at which the dropping edge passes half the maximum number of hits 
in the histogram. This value was measured in bins along the 200~mm RTPC's 
length to take into account variations in the electric and magnetic field in 
the RTPC (see Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

@@

Due to the non perfect experimental conditions, in particular possible contamination 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around 
run number 61600. While our gas system was kept slightly over atmospheric 
pressure to limit contamination from air or other external gases, it is likely 
that this leak was the source of modification of the drift speed.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for individual 
runs (approximately 2 hours long). We observe significant change in the drift 
speed before and after run 61600, while variations within these periods are 
around 2\%.

@@

In summary, we obtain from our calibration a parametrization of the drift speed 
as a function of both position along the beam axis and the run number.  These 
functions were extracted for our entire data set and implemented in the track 
reconstruction code.

\subsection{Drift Path Calibration}

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{Biagi:1999nwa}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used this calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometrical accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{Fenker:2008zz}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) was extracted as a function of time.  In 
Figure~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one bin, 
where the drift path is easily identified and eventually fitted for 
implementation in our reconstruction code.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Figure~\ref{fig:Drift_run_number_1}). Interestingly, we found very similar 
drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

To calibrate the gains, we compared the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charge along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Figure~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated as the 
ratio of the measured ADCs to the simulated deposited energy.  Then, these 
gains were refined using correction factors obtained from a sample of good 
tracks. For each track, we calculate the corrected energy deposit on a pad and 
compared it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Figure~\ref{fig:dedx_p_exp_2nd}, where energy loss 
of particles is plotted against momentum over charge ratio. One can clearly see 
the band for $^4$He in its expected position.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Figure~\ref{fig:noise} for a particularly noisy 
channel.  Its amplitude is similar to those of real tracks. About 18\% of the 
readout channels exhibit large contributions from this noise characteristic.  
Due to its unique time-energy correlation for the given channels, the noise 
could be removed on an event by event and channel by channel basis without 
significant loss of good signals. The result of the
procedure is illustrated in the bottom panel of Figure~\ref{fig:noise}.

@@

The primary data sample used for calibration and performance assessment of the
RTPC was elastic scattering with a 1.2 GeV electron beam. The electron momentum
and direction is measured with CLAS, which uniquely determines the expected recoiling
$^4$He kinematics. Matching requirements between reconstructed and expected $z$-vertex
and direction of the RTPC track provides a clean selection of elastically-scattered
$^4$He, shown in Figure~\ref{fig:w}.

\begin{figure}[tb]\centering
  \includegraphics[width=8cm]{fit_W_distribution_l.png}
  \caption{The recoil mass $W$ distribution calculated from electron kinematics 
  before (``inclusive'') and after (``exclusive'') requiring a matching track 
  in the RTPC.\label{fig:w}}
\end{figure}
